Estimating duration using survival analysis

### 2024-07-29 cis-perfect-testing.tex
Rob p84 comment
Rob first p87 comment
Rob third p87 comment

The first complication in the analysis is that infection episodes can be undetected.
Undetected infection episodes are episodes with no associated positive tests.
Therefore, there is no information directly observed about the episode; even the number that occur is latent.
They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
Undetected episodes are, on average, shorter than detected episodes; therefore, neglecting them would lead to a biased estimate of the survival distribution.
To see why they are shorter, consider the infection episode in \cref{perf-test:fig:truncation}.
Here two individuals are shown, with identical data collected for each of them (a series of negative
 tests).
The first individual was never infected whereas the second individual has an undetected infection. 
The undetected infection episode begins at time 33 and has a 10-day duration.
The individual would have been detected at their next test if their duration had instead been 24 days or more.
The gap between tests in the CIS (up to 28 days, see \cref{E-intro:sec:cis}) is longer than the duration of most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
Therefore, it is likely that many infections in the CIS cohort were undetected.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{%
    Identical data for two individuals, with a series of negative tests.
    Top individual remains uninfected, bottom individual has an undetected infection.
    The shaded area represents when the individual is detectable.
  }
  \label{perf-test:fig:truncation}
\end{figure}

@@

The second complication is that the beginning and end times of detected infection episodes are latent.
Recall, from \cref{E-biology-data:sec:cis-episodes}, that the time episode $j$ starts, $b_j$, is only known to lie in the interval $[l_j^{(b)}, r_j^{(b)}]$ where $l_j^{(b)}$ is the day after the negative test prior to the start of episode $j$ and $r_j^{(b)}$ is the first positive test in episode $j$.
Similarly, for the end of the infection episode, $e_j \in [l_j^{(e)}, r_j^{(e)}]$.
\Cref{perf-test:fig:double-interval-censor} shows this issue graphically.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=\textwidth]{cis-perfect-testing/double-interval-censor}}
  \caption[Double-interval censoring in CIS data]{%
        The beginning and end of each episode observed in the CIS data are only known up to an interval.
        Demonstrated here by a participant who is recorded negative at time 7 and positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode.
    }
    \label{perf-test:fig:double-interval-censor}
\end{figure}

@@

I analyse the $\ndet = \numprint{4800}$ episodes with their first positive test between 10 Oct 2020 and 6 Dec 2020 inclusive, with negatives bounding the start and end time of the episode; I refer to these episodes as \emph{detected}; in \cref{perf-test:sec:prob-undetected} I formalise the conditions for an episode to be detected.
Each day in this period is denoted as $t = 1, \dots, T$.
I use these episodes because a negative test before the start of the episode provides a lower bound on the episode's start time; otherwise, there is little information on its length.
Similar information comes from the negative test after the end of the episode, although a negligible number of episodes are missing a negative test after the end of the episode.
How the episode data are constructed is detailed in \cref{E-biology-data:sec:cis-episodes}.

@@

For any episode $j$, define $O(j)$ as the observations of that episode.
Test results are deterministic based on the time of the test and the individual's infection status.
Therefore, $O(j)$ is fully determined by the triplet $(b_j, e_j, i(j))$ consisting of $j$'s beginning time, end time, and the individual in which the episode occurs respectively.
The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).

@@

If $j$ is undetected, there are no observations; in which case, define $O(j) = \emptyset$.
Let $\nnodet$ denote the (latent) number of undetected episodes experienced across all $\Ncis$ individuals.
Therefore, the total number of infection episodes in the cohort is $\ntot = \ndet + \nnodet$ and I index them with $j = 1, \dots, \ntot$.

>>>>

The first complication in the analysis is that infection episodes can be undetected.
Undetected infection episodes are episodes with no associated positive tests.
Therefore, there is no information directly observed about the episode; even the number that occur is latent.
They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
Undetected episodes are, on average, shorter than detected episodes; therefore, neglecting them would lead to a biased estimate of the survival distribution.
To see why they are shorter, consider the infection episode in \cref{perf-test:fig:truncation}.
Here two individuals are shown, with identical data collected for each of them (a series of negative
 tests).
The first individual was never infected whereas the second individual has an undetected infection. 
The undetected infection episode begins at time 33 and has a 10-day duration.
The individual would have been detected at their next test if their duration had instead been 24 days or more.
The gap between tests in the CIS (up to 28 days, see \cref{E-intro:sec:cis}) is longer than the duration of most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
Therefore, it is likely that many infections in the CIS cohort were undetected.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{%
    Identical data for two individuals, with a series of negative tests.
    Top individual remains uninfected, bottom individual has an undetected infection.
    The shaded area represents when the individual is detectable.
    Time is relative to the individual's first CIS test.
  }
  \label{perf-test:fig:truncation}
\end{figure}

@@

The second complication is that the beginning and end times of detected infection episodes are latent.
Recall, from \cref{E-biology-data:sec:cis-episodes}, that the time episode $j$ starts, $b_j$, is only known to lie in the interval $[l_j^{(b)}, r_j^{(b)}]$ where $l_j^{(b)}$ is the day after the negative test prior to the start of episode $j$ and $r_j^{(b)}$ is the first positive test in episode $j$.
Similarly, for the end of the infection episode, $e_j \in [l_j^{(e)}, r_j^{(e)}]$.
\Cref{perf-test:fig:double-interval-censor} shows this issue graphically.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=\textwidth]{cis-perfect-testing/double-interval-censor}}
  \caption[Double-interval censoring in CIS data]{%
        The beginning and end of each episode observed in the CIS data are only known up to an interval.
        Demonstrated here by a participant who is recorded negative at time 7 and positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode.
      Time is relative to the individual's first CIS test.
    }
    \label{perf-test:fig:double-interval-censor}
\end{figure}

@@

I analyse the $\ndet = \numprint{4800}$ episodes with their first positive test between 10 Oct 2020 and 6 Dec 2020 inclusive, with negatives bounding the start and end time of the episode; I refer to these episodes as \emph{detected}; in \cref{perf-test:sec:prob-undetected} I formalise the conditions for an episode to be detected.
Each day in this period is denoted as $t = 1, \dots, T$; \ie day $1$ is the 10 Oct 2020 and day $T$ is the 6 Dec 2020.
I use these episodes because a negative test before the start of the episode provides a lower bound on the episode's start time; otherwise, there is little information on its length.
Similar information comes from the negative test after the end of the episode, although a negligible number of episodes are missing a negative test after the end of the episode.
How the episode data are constructed is detailed in \cref{E-biology-data:sec:cis-episodes}.

@@

For any episode $j$, define $O(j)$ as the observations of that episode; $j = 1, \dots, \ntot$ where $\ntot$ is the total number of infection episodes in the cohort.
Test results are deterministic based on the time of the test and the individual's infection status.
Therefore, $O(j)$ is fully determined by the triplet $(b_j, e_j, i(j))$ consisting of $j$'s beginning time, end time, and the individual in which the episode occurs respectively.
The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).

@@

If $j$ is undetected, there are no observations; in which case, define $O(j) = \emptyset$.
Let $\nnodet$ denote the (latent) number of undetected episodes experienced across all $\Ncis$ individuals.
Therefore, the total number of infection episodes in the cohort is $\ntot = \ndet + \nnodet$.
